Patterns of stress tolerance versus resistance in genome-wide expression data

### 2014-07-25 ApTranscriptome_TR.Rmd
Updates. Change TR to download expression data rather than re-run sailfish as unclear if filtered fastq files will be archived.

In this technical report, which accompanies the manuscript **Thermal reactionome of a common ant species** (Stanton-Geddes et al., in press), we:

1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. evaluate differences in the expression patterns between the two species
3. perform gene set enrichment analysis of thermally-responsive genes for the two species

@@

The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.

@@

The Illumina fastq files are available from [https://minilims1.uvm.edu/BCProject-26-Cahan/downloads.html]. The Trimmomatic filtered fastq files should be downloaded, uncompressed and moved to the appropriate directory using the following code.

~~~
# download filtered fastq files, uncompress and move
wget --no-check-certificate https://minilims1.uvm.edu/BCProject-26-Cahan/_downloads/trimmomatic_output.tar.gz
tar -zxvf trimmomatic_output.tar.gz
mkdir -p data/
mv ind_files data/.
~~~

@@

The Illumina reads were filtered using the program [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) `r citep("10.1093/nar/gks540")` to remove Ilumina adapter sequences and filter out bases with quality scores less than ??.

~~~
TRIMMOMATIC CODE
~~~

This filtering yielded...

These reads were combined and used in *de novo* transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`. Note that this required ??? GB RAM and ??? hours run-time and was run on the [Vermont Genetics Network](http://vgn.uvm.edu/) computing cluster.

~~~
TRINITY CODE
~~~

@@

Then build the index of the assembly:

```{r sailfish_index}
system("sailfish index -t results/trinity-full/Trinity_cap3_uclust.fasta -o results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4")
```

Once this is done, quantify expression for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are three four filtered read files:

@@

Make a directory for the expression values

```{r mk_sailfish_dir}
system("mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust")
```

Then, for each sample, run the following command in the `results/trinity-full/` directory:

    sailfish quant -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=SE:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq -p 4

@@

Or, using a loop in R:

```{r sailfish_quant, results = 'hide'}
# directory containing trimmed reads
readdir <- "data/ind_files/" 
# sailfish index directory
sfindex <- "results/trinity-full/sailfish-index-Trinity-cap3-uclust"
# sailfish expression directory
sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"

# list of reads in each of four trimmed classes
readlist <- list.files(readdir)
paired.left <- readlist[grep(".\\.paired.left.fastq$", readlist)]
paired.right <- readlist[grep("\\.paired.right.fastq$", readlist)]
unpaired.left <- readlist[grep("unpaired.left.fastq$", readlist)]
unpaired.right <- readlist[grep("unpaired.right.fastq$", readlist)]

# Loop across each sample and quantify expression

# NOTE - samples listed in same order as given by the above lists
samples <- c("A22-0", "A22-10", "A22-14", "A22-17", "A22-21", "A22-24", "A22-28", "A22-31", "A22-35", "A22-38", "A22-3", "A22-7", "Ar-0", "Ar-10", "Ar-14", "Ar-17", "Ar-21", "Ar-24", "Ar-28", "Ar-31", "Ar-35", "Ar-38", "Ar-3", "Ar-7")

for (j in 1:length(samples)) {
    message("Start expression quantification for sample ", samples[j], ": ", Sys.time())
    quantdir <- paste(sfexpressionroot, samples[j], "_quant", sep="")
    samp.pos <- grep(paste(paste(samples[j], "_", sep="")), paired.left)
    samp.paired.l <- paste(readdir, paired.left[samp.pos], sep="")
    samp.paired.r <- paste(readdir, paired.right[samp.pos], sep="")
    samp.unpaired.l <- paste(readdir, unpaired.left[samp.pos], sep="")
    samp.unpaired.r <- paste(readdir, unpaired.right[samp.pos], sep="")
    sailfishcmd <- paste("sailfish quant -i ", sfindex, " -o ", quantdir, " -l 'T=SE:S=U' -r ", samp.paired.l, " ", samp.paired.r, " ", samp.unpaired.l, " ", samp.unpaired.r, " -p 4", sep="")
    #print(sailfishcmd)
    system(sailfishcmd)
    message("Done with expression quantification for sample ", samples[j], ": ", Sys.time(), "\n")
}
```

This generated a directory for each sample

`r list.files(sfexpressionroot)`

and within each directory there are the following r:

`r list.files(quantdir)`

The file *quant_bias_corrected.sf* contains the following columns, following a number of header lines:

@@

```{r load_expression_data, eval=TRUE, echo=FALSE, results='hide'}
# read in each file using loop
for (j in 1:length(samples)) {
    samp <- samples[j]
    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
    outpre <- gsub("-", "_", samp)
    outname <- paste(outpre, "_quant", sep="")
    read.sailfish.quant(filein=paste(sfexpressionroot, samp, "_quant/quant_bias_corrected.sf", sep=""), outname=outname, samp = samp, trtval = trtval)
}

>>>>

In this technical report, which accompanies the manuscript **Patterns of stress tolerance versus resistance in genome-wide expression data of parapatric ant species** (Stanton-Geddes et al., in press), we:

1. describe the *de novo* assembly of the transcriptome for two ant species within the *Aphaenogaster rudis-picea-fulva* species complex `r citep("10.1155/2012/752815")`
2. identify thermally-responsive genes
3. evaluate differences in reaction norms of expression for each thermally-responsive gene between the two species
4. perform gene set enrichment analysis of thermally-responsive genes for the two species

@@

@@

The raw Illumina fastq files are available from the NCBI short read archive [link tbd]. The assembled transcriptome, annotation and expression values are downloaded rather than re-run due to the computational demands, but the exact commands for each of these steps are documented below.

@@

The Illumina reads were filtered using the program [Trimmomatic](http://www.usadellab.org/cms/?page=trimmomatic) `r citep("10.1093/nar/gks540")` to remove Ilumina adapter sequences, trim bases with PHRED quality scores less than 15 in a 4 bp sliding window and remove final trimmed sequences with length less than 36 bp. The code used was

~~~
java -jar trimmomatic-0.30.jar PE -threads 40 -phred33 -trimlog trimmomatic.log sample.R1.fastq sample.R2.fastq sample.R1.trimmed.paired.fastq sample.R1.trimmed.unpaired.fastq sample.R2.trimmed.paired.fastq sample.R2.trimmed.unpaired.fastq ILLUMINACLIP:TruSeq3-PE.fa:2:30:10 LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:36
~~~

where "sample" was replaced by each sample. This filtering yielded 339,845,787 properly paired reads and 16,860,885 unpaired reads.

Properly paired and unpaired reads passing the Trimmomatic filter were combined and used in de novo transcriptome assembly using the program [Trinity](http://trinityrnaseq.sourceforge.net/) `r citep("10.1038/nbt.1883")`. Unpaired reads resulting from Trimmomatic were concatenated to the set of left (R1) reads per Trinity usage documentation. Assembly on a single 40 CPU machine with 1TB of memory required 20 hours of compute time.

~~~
Trinity.pl --seqType=fq --JM=90G —left=all.R1.trimmed.fastq —right=all.R2.trimmed.fastq --output=trinity --CPU=40 --inchworm_cpu=40 --bflyCPU=5
~~~

@@

An index of the assembly is built with the command:

~~~
sailfish index -t results/trinity-full/Trinity_cap3_uclust.fasta -o results/trinity-full/sailfish-index-Trinity-cap3-uclust -k 20 -p 4
~~~

Once this is done, expression is quantified for the Trimmomatic filtered reads from each species-treatment sample separately. Note that for each sample, there are four filtered read files:

@@

~~~
# make a directory for the expression values
mkdir -p results/trinity-full/sailfish-expression-Trinity-cap3-uclust
# change to "trinity-full" directory
cd results/trinity-full
# for each sample, run the following command
sailfish quant -i sailfish-index-Trinity-cap3-uclust -o sailfish-expression-Trinity-cap3-uclust/A22-0 -l "T=SE:S=U" -r A22-0_ATCACG.unpaired.left.fastq A22-0_ATCACG.unpaired.right.fastq A22-0_ATCACG.paired.left.fastq A22-0_ATCACG.paired.right.fastq -p 4
~~~

@@

These files are downloaded for convenience:

~~~
wget ...
mkdir -p /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
mv .. /results/trinity-full/sailfish-expression-Trinity-cap3-uclust
tar -zxvf ...
~~~

For each sample, there is a directory containting a file *quant_bias_corrected.sf*. This file has the following columns, following a number of header lines:

@@

```{r load_expression_data, eval=TRUE, echo=FALSE, results='hide'}
# sailfish expression directory
sfexpressionroot <- "results/trinity-full/sailfish-expression-Trinity-cap3-uclust/"

# read in each file using loop
for (j in 1:length(samples)) {
    samp <- samples[j]
    trtval <- as.numeric(str_split_fixed(samp, "-", 2)[2])
    outpre <- gsub("-", "_", samp)
    outname <- paste(outpre, "_quant", sep="")
    read.sailfish.quant(filein=paste(sfexpressionroot, samp, "_quant/quant_bias_corrected.sf", sep=""), outname=outname, samp = samp, trtval = trtval)
}

### 2014-07-25 doc/ApTranscriptome_manuscript.Rmd
Updates. Change TR to download expression data rather than re-run sailfish as unclear if filtered fastq files will be archived.

The patterns of gene expression between species differed substantially, with more down-regulation in *A. carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A. picea* in response to thermal extremes (**High** and **Low** genes; Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are up-regulated at cold temperatures (**Low**) in *A. picea* but shut-down **Intermediate** by *A. carolinensis* include metabolism, transport and regulation GO clusters. Thus, *A. carolinensis* shuts-down metabolism to a lesser extent than *A. picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis*, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically-enduced resistance.

>>>>

The patterns of gene expression between species differed substantially, with more down-regulation in *A. carolinensis* (**Intermediate** genes) and more up-regulation of gene expression in *A. picea* in response to thermal extremes (**High** and **Low** genes; Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this pattern. The genes that are up-regulated at cold temperatures (**Low**) in *A. picea* but shut-down **Intermediate** by *A. carolinensis* include metabolism, transport and regulation GO clusters. Thus, *A. carolinensis* shuts-down metabolism to a lesser extent than *A. picea* and relies on other mechanisms such as behavior or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis*, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we find that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result is consistent with work in corals [@barshis2013] that shows that organisms in more thermally-stressful environments are more likely to invest in a tolerance strategy, while organisms from less stressful environments rely on plastically induced resistance.
